Fail Faster

### 2025-03-22 paper/paper.tex
rough stab

For STLC, staging accounts for the bulk of performance gains, yielding
$1.73-4.05\times$ speedups for the type-derived STLC generator and $4.10-5.71\times$ for
the well-typed generator. Adding \csplitmix, these numbers increase to
$2.90-5.39\times$ and $5.33-8.33\times$.

@@

The variation in speedups raises a natural question: what 
determines how much a generator benefits from staging? One 
likely factor is its degree of reliance on generator library functions.
This is hard to quantify directly, since not all functions 
contribute equally to overhead. As a proxy, we count how many 
times a generator calls \texttt{bind}---the central function in 
monadic generator libraries, extensively used both directly and within 
other combinators.

We select four benchmarks with disparate speedups: insertion-based BSTs, single-pass BSTs, 
boolean lists, and well-typed STLC terms. For each generator, 
we run 10,000 trials using a fixed generation size of 100, and 
record the average number of \texttt{bind} calls. Results in \figref{fig:binds} show a clear linear 
correlation between the number of \texttt{bind} calls and the speedup 
achieved by staging, supporting the idea that generators which make heavier use of 
generator library functions tend to benefit more.

@@

Similarly, we investigate the impact of \csplitmix by tracking the number of calls to \texttt{next\_int64} in \bq's \texttt{splittable\_random} 
library, which samples the next random integer.
This, too, is an imperfect proxy, as the SplitMix algorithm's runtime does plenty of stuff that isn't 
sampling which can dominate its runtime---which is all faster in \csplitmix. However, it too validates our intuition that sampling-heavy generators achieve
greater benefit from \csplitmix than they do staging:

>>>>

For STLC, staging accounts for larger performance gains, yielding
$1.73-4.05\times$ speedups for the type-derived STLC generator and $4.10-5.71\times$ for
the well-typed generator. Adding \csplitmix, these numbers increase to
$2.90-5.39\times$ and $5.33-8.33\times$.

@@

\subsubsection{Teasing apart speedups} \tr{Working title...}

The variation in speedups in \secref{subsubsection:ocaml} raises a natural question: what 
determines how much a generator benefits from staging, \rand optimizations, or both?
One likely explanation is that the two techniques address different performance bottlenecks:
generators that sample more often benefit more from improved random sampling, while those that rely
heavily on generator library combinators gain more from staging.
To test this idea, we look for measurable proxies that approximate a generator's reliance on its generator and randomness libraries.
For the former, we use the number of calls to \texttt{bind}---the central monadic generator function, extensively used both directly and within other combinators.
For the latter, we count the number of times a generator invokes the sampling function \texttt{next\_int64}. We select four strategies from \secref{subsubsection:ocaml} with disparate speedups: insertion-based BSTs, single-pass BSTs, 
boolean lists, and well-typed STLC terms.

To determine how often \texttt{bind} is invoked, we generate 10,000 values using a fixed generation size of 100, and 
record the average number of \texttt{bind} calls. Results in \figref{fig:binds} show a clear linear 
correlation between the number of \texttt{bind} calls and the speedup 
achieved by staging, supporting the idea that generators that make heavier use of 
generator library functions benefit more.

@@

To determine the number of random samples, we repeat the same test, but use Intel ProcessorTrace to capture processor activity.
We then count the number of invocations of \texttt{next\_int64} and average this over the number of complete generations in the trace. As expected,
the BST strategies, which achieve large speedups from \csplitmix, invoke the sampling function more: 353 times for the insertion-based strategy, and 271 for the single-pass strategy. 
STLC, which sees moderate improvements from \csplitmix, invokes \texttt{next\_int64} 211 times, and the boolean list strategy, which sees no improvement from \csplitmix, invokes it only 100 times.
